Epidemiological birth-death models with partner notification

### 2024-06-10 article/article_plos.tex
PLOS formatted article

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,mathrsfs}\usepackage{amsmath,amsfonts,amssymb,float,graphicx,mathtools,mathrsfs,tikz,amsthm}

\usepackage{standalone}
\usepackage{caption}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usepackage{tikzsymbols}
\usepackage{hieroglf}
\usepackage{nameref}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{stackengine}

@@

%opening
\title{Epidemiological birth-death models with partner notification}
\author{Anna Zhukova et al.}

\begin{document}

\maketitle

%\begin{abstract}
%
%\end{abstract}

\section{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\citep{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e. genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\citep{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

Phylodynamic models can be classified into two main families:  coalescent~\citep{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\citep{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\citep{Macpherson2021}. In the classic BD model with incomplete sampling~\citep{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g. due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

@@

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by~\citet{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g. SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g. Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g. HIV) more than the others~\citep{Stadler2014}.

To allow for heterogeneity over time, \citet{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them. \citet{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types. In particular the skyline approach can be useful to account for changes in the sampling policies (e.g. before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).

Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling. We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for correlation effects in sampling, in particular due to partner notification (contact tracing). Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection. It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\citep{el-sadrContactTracingBarriers2022}. For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\citep{worldhealthorganizationCountryPolicyReview2016}.

@@

\section{BD-PN Framework}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
while bifurcations (i.e. internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}). The tree branches are measured in units of time. Let $T$ denote the time that passed between the tree root (the beginning of the (sub-)epidemic, at time $t=0$) and the last sampled tip (at $t=T$).

\begin{figure*}[tbhp]
\centering 
%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
\includestandalone[width=0.6\textwidth]{Fig_tree}
\caption{A transmission tree $\mathscr{T}$ with $n=5$ external nodes (i.e. tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g. $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
\label{fig:tt} 
\end{figure*}

\subsection{BD Model with incomplete sampling}
Under the basic birth-death model~\citep{Stadler2009} all the individuals are in the same state. They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, and their pathogen can be sampled upon removal with a constant probability $\rho$.

@@

\subsection{MTBD models}
MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes. A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ ($k \neq l$), and transmit their pathogen to an individual in state $r \in \{1, \ldots, m\} $ at a constant average rate $\lambda_{kr}$. The time between events % of the same type 
is hence modeled with exponential distributions.

\subsection{Partner Notification extension}

@@

These assumptions can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\citep{Voznica2021}, which does not require likelihood calculation).

@@

\subsection{MTBD-PN tree simulator}

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in Appendix. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.

\section{Testing for presence of Partner Notification}~\label{sec:test}
To access whether a given tree reveals the presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).

@@

More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models). In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar. Therefore, in the absence of partner notification two external branches taken randomly from the tree and two external branches belonging to the same cherry should be indistinguishable, if they all start at about the same time. Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$. Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\citep{mooersBranchLengthsBirth2012}. %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.

@@

\begin{figure*}[tbhp]
\centering 
\includegraphics[width=0.8\textwidth]{Fig_cherryswap.png}
\caption{Reshuffled cherry generation. \textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left). The randomly selected tip is shown with a bold branch for each cherry (e.g. the tip on the right for the leftmost cherry). The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$). \textbf{b.} Reshuffled cherries, which were obtained from the original ones by swapping the selected tips between the neighbouring cherry couples (as shown with arrows in a.): the first and the second; the third and the forth. As the total number of cherries (7) is odd, the last three (instead of two) cherries exchanged their selected tips in a cycle. The reshuffled tip sampling time differences are shown for the two leftmost cherries ($d'_1$ and $d'_2$).}
\label{fig:tipswap} 
\end{figure*}

@@

We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~\ref{sec:sim}).

\subsection{Performance on simulated data}~\label{sec:test:sim}
To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models. (See Fig.~1 in \citet{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).

For each model, we generated 100 trees with 500--1\,000 tips with our MTBD/MTBD-PN tree simulator. For each tree the parameter values were drawn randomly within the following bounds:
\begin{itemize}
\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
\item sampling probability $\rho \in ]0.1, 0.9]$,
\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
\end{itemize}

We then run the PN tests on each tree of these 6 datasets. For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e. 98\%/97\%/95\% true negative and 2\%/3\%/5\% false positive results). For 95/95/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e. 95\%/95\%/96\% true positive and 5\%/5\%/4\% false negative results). These results are summarised in Table~\ref{tbl:pntest}. Our test therefore showed both high specificity and sensitivity.

\section{Mathematical formulation of BD and BD-PN models}\label{math}

\subsection{Master equation and tree likelihood under BD model}
The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without this individual nor any of their induced cases being sampled:

@@

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$. The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree. $t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch). % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

@@

Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}). To become identifiable it requires one of its parameters to be fixed. In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).

\subsection{Master equation and tree likelihood under BD-PN model}

\subsubsection{Tree branches under BD vs BD-PN models}

@@

This would lead to a shorter oriented part on the B branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).

>>>>

% Template for PLoS
% Version 3.6 Aug 2022
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs,amsfonts,mathtools,mathrsfs,amsthm}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage[nopatch=eqnum]{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

\usepackage{standalone}
\usepackage{tikz,tikzsymbols}
\usetikzlibrary{positioning}
\usepackage{booktabs}
\usepackage{stackengine}

@@

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Epidemiological birth-death models with partner notification} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anna Zhukova\textsuperscript{1,2,3*},
Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
Olivier Gascuel\textsuperscript{1,4},
\\
\bigskip
\textbf{1} Unit\'e Bioinformatique Evolutive, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{2} Bioinformatics and Biostatistics Hub, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{3} G5 Evolutionary Dynamics of Infectious Diseases, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{4} Institut de Syst\'{e}matique, Evolution, Biodiversit\'{e} (ISYEB) - URM 7205 CNRS, Mus\'{e}um National d'Histoire Naturelle, SU, EPHE \& UA, Paris, France
\\
\bigskip

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* anna.zhukova@pasteur.fr

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e. genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g. due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

@@

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g. SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g. Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g. HIV) more than the others~\cite{Stadler2014}.

To allow for heterogeneity over time, Stadler \textit{et al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them. Kühnert \textit{et al.}~{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types. In particular the skyline approach can be useful to account for changes in the sampling policies (e.g. before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).

Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling. We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for correlation effects in sampling, in particular due to partner notification (contact tracing). Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection. It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}. For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.

@@

\section*{Materials and methods}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
while bifurcations (i.e. internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}). The tree branches are measured in units of time. Let $T$ denote the time that passed between the tree root (the beginning of the (sub-)epidemic, at time $t=0$) and the last sampled tip (at $t=T$).

\begin{figure}[!h]
\centering 
%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
\includestandalone[width=0.6\textwidth]{Fig_tree}
\caption{\textbf{Transmission tree} $\mathscr{T}$ with $n=5$ external nodes (i.e. tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g. $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
\label{fig:tt} 
\end{figure}

\subsection*{Extending Multi-Type Birth-Death models with Partner Notification}
\subsubsection*{Birth-Death (BD) model with incomplete sampling}
Under the basic birth-death (BD) model~\cite{Stadler2009} all the individuals are in the same state. They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, and their pathogen can be sampled upon removal with a constant probability $\rho$.

@@

\subsubsection*{Multi-Type Birth-Death (MTBD) models}
Multi-Type Birth-Death (MTBD) models~\cite{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes. A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ ($k \neq l$), and transmit their pathogen to an individual in state $r \in \{1, \ldots, m\} $ at a constant average rate $\lambda_{kr}$. The time between events % of the same type 
is hence modeled with exponential distributions.

\subsubsection*{Partner Notification (PN) extension}

@@

These assumptions can be easily relaxed in a tree simulator (e.g., to be used with a deep learning estimator~\cite{Voznica2021}, which does not require likelihood calculation).

@@

\subsection*{MTBD-PN tree simulator and test datasets}\label{sec:sim}

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in~\nameref{S1_Appendix}. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.

Using our simulator, we generated trees under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\cite{Stadler2014}) and BDSS (Birth-Death with Super Spreading~\cite{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models. (See Fig.~1 in \cite{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).

For each model, we generated 100 trees with 500--1\,000 tips. For each tree the parameter values were drawn randomly within the following bounds:
\begin{itemize}
\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
\item sampling probability $\rho \in ]0.1, 0.9]$,
\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
\end{itemize}

\subsection*{Testing for presence of Partner Notification}~\label{sec:test}
To access whether a given tree reveals the presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).

@@

More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models). In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar. Therefore, in the absence of partner notification two external branches taken randomly from the tree and two external branches belonging to the same cherry should be indistinguishable, if they all start at about the same time. Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$. Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\cite{mooersBranchLengthsBirth2012}. %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.

@@

\begin{figure*}[tbhp]
\centering 
\includegraphics[width=0.5\textwidth]{Fig_cherryswap.png}
\caption{Reshuffled cherry generation. \textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left). The randomly selected tip is shown with a bold branch for each cherry (e.g. the tip on the right for the leftmost cherry). The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$). \textbf{b.} Reshuffled cherries, which were obtained from the original ones by swapping the selected tips between the neighbouring cherry couples (as shown with arrows in a.): the first and the second; the third and the forth. As the total number of cherries (7) is odd, the last three (instead of two) cherries exchanged their selected tips in a cycle. The reshuffled tip sampling time differences are shown for the two leftmost cherries ($d'_1$ and $d'_2$).}
\label{fig:tipswap} 
\end{figure*}

@@

We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~"\nameref{sec:sim}").

\subsection*{Mathematical formulation of BD and BD-PN models}\label{math}

\subsubsection*{Master equation and tree likelihood under BD model}
The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without this individual nor any of their induced cases being sampled:

@@

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$. The master equations for the BD models were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree. $t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch). % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

@@

Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\cite{Stadler2009}). To become identifiable it requires one of its parameters to be fixed. In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).

\subsubsection*{Master equation under BD-PN model}

\paragraph{Tree branches under BD vs BD-PN models.}

@@

This would lead to a shorter partner-before-notification part on the B branch (finishing at $t_C$), and a longer notified-partner part (between $t_C$ and $t_B$).
